On Extending a Full-Sharing Multithreaded Tabling Design with Batched Scheduling

### 2015-05-27 paper.tex
paper -> revising 1 -> end for noew

\begin{abstract}
Tabling is an implementation technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Multithreading in Prolog is the
ability to concurrently perform computations, in which each
computation runs independently but shares the program clauses. When
multithreading is combined with tabling, we have the best of both
worlds, since we can exploit the combination of higher procedural
control with higher declarative semantics. To support this
combination, the Yap Prolog system has, at engine level, multiple
designs that vary from a No-Sharing design, where each thread
allocates fully private tables for each new tabled subgoal called
during its computation, to a Full-Sharing design, where threads share
the complete table space. Arguably, batched scheduling is one of the
most successful tabling scheduling strategies. In this work, we
propose the PAC strategy to combine the Full-Sharing design with
batched scheduling. Experimental results show that, PAC is a good
first approach, since with it the FS design remains quite
competitive.\\

@@

Tabling~\cite{Chen-96} is a technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Tabling consists in storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of XSB. The
success of SLG-WAM led to several alternative implementations that
differ in the execution rule, in the data-structures used to implement
tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

Multithreading in Prolog is the ability to concurrently perform
computations, in which each computation runs independently but shares
the program clauses. When multithreading is combined with tabling, we
have the best of both worlds, since we can exploit the combination of
higher procedural control with higher declarative semantics. To the
best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
are the only Prolog systems that support the combination of
multithreading with tabling. Yap implements a SWI-Prolog compatible
multithreading library~\cite{Wielemaker-03}. Yap's threads have their
own execution stacks and only share the code area where predicates,
records, flags and other global non-backtrackable data are stored. For
multithreaded tabled evaluation, a thread views its tables as private
but, at the engine level, Yap has three
designs~\cite{Areias-12a}. These designs vary from a \emph{No-Sharing}
(NS) design, where each thread allocates private tables for each new
tabled subgoal called during its evaluation, to a \emph{Full-Sharing}
(FS) design, where threads share the complete table space.

The decision about the evaluation flow is determined by the
\emph{scheduling strategy}. Different strategies may have a
significant impact on performance, and may lead to a different
ordering of solutions to the query goal. Arguably, the two most
successful tabling scheduling strategies are \emph{local scheduling}
and \emph{batched scheduling}~\cite{Freire-96}. Local scheduling tries
to complete subgoals as soon as possible. When new answers are found,
they are added to the table space and the evaluation fails. Answers
are only returned when all program clauses for the subgoal at hand
were resolved, i.e. after the complete evaluation of subgoal's search
space. Local scheduling has the advantage of minimizing the size of
clusters of dependent subgoals, however it delays propagation of
answers and requires the complete evaluation of the search
space.

@@

In this work, we propose an extension to the table space data
structures, which we named \emph{Private Answer Chaining (PAC)}, as a
way to keep track, per thread and subgoal call, of the answers that
were already found and already propagated. We discuss in detail our
proposal for extending the FS design with batched scheduling and we
present a performance analysis comparison between local and batched
scheduling. Experimental results show that, despite the extra PAC data
structures required to support batched scheduling with the FS design,
the execution time of the combination is still quite competitive when
comparing against the FS design with local scheduling.

The remainder of the paper is organized as follows. First, we briefly
introduce some background and related work. Then, we describe our
approach for combining batched scheduling with the FS design and we
discuss the most important implementation details. Finally, we present
experimental results and we end by outlining some conclusions.

@@

\begin{figure}[!ht]
\vspace{-0.5\intextsep}
\centering
\includegraphics[width=9.5cm]{figures/yap-mt.pdf}
\caption{Yap's multithreaded table space organization for the NS and FS designs}
\label{fig_yap_mt_support}
\vspace{-1.0\intextsep}
\end{figure}

@@

On the other hand, batched scheduling schedules the evaluation of a
program in a depth-first manner. It favors the forward execution first
instead of backtracking, leaving the consumption of answers and
completion for last. It thus tries to delay the need to move around
the search tree by batching the return of answers. When new answers
are found for a particular tabled subgoal, they are added to the table
space and the execution continues. For some situations, this results
in creating dependencies to older subgoals, therefore enlarging the
current CDS and delaying the fix-point that guarantees that all
dependent subgoals in a CDS are completely
evaluated~\cite{Sagonas-98}. Batched scheduling can be an useful
strategy in tabled logic programs that require an eager propagation of
answers and/or do not require the complete set of answers to be found
(partial evaluation of the search space).

@@

\begin{wrapfigure}{r}{6.5cm}
\vspace{-\intextsep}
\centering
\includegraphics[width=6.5cm]{figures/pcc_key-idea.pdf}
\caption{PAC splits answer representation and answer propagation}
\label{fig_tabtries_pcc_key-idea}
\vspace{-0.5\intextsep}
\end{wrapfigure}

In this section, we describe our proposal to support the combination
of batched scheduling with the FS design. Remember that batched
scheduling requires further support to work with the FS design
because, with batched scheduling, we have to detect when a newly found
answer is also a new answer for the thread at hand in order to ensure
that new answers for a given subgoal call are correctly propagated one
and only once to all thread’s repeated calls. In the FS design answer
representation and answer propagation is stored in the answer trie
data structure, thus threads are unable to know if they have or not
have propagated an answer whenever it appears during the
evaluation. To solve that, we propose an extension to the table space
data structures, which we named \emph{Private Answer Chaining} (PAC),
as a way to keep track, per thread and subgoal call, of the answers
that were already found and propagated to the thread's repeated
calls. Figure~\ref{fig_tabtries_pcc_key-idea} illustrates the key idea
of the PAC strategy. In a nutshell, PAC splits answer representation
from answer propagation, and allows the first to be kept publicly
shared among threads in the answer trie data structure, and the second
to be privately stored in the subgoal frame data structure of each
thread. Thus, with PAC, whenever a answer appears during the
evaluation, a thread knows if the answer was already propagated or not
propagated, just by consulting its own subgoal frame data structure.

%% \begin{figure}[!ht]
%% \centering
%% \includegraphics[width=7.5cm]{figures/pcc_key-idea.pdf}
%% \caption{PAC splits answer representation and answer propagation}
%% \label{fig_tabtries_pcc_key-idea}
%% \end{figure}

@@

The PAC procedure works at the subgoal frame level, which is private
to each thread. The key idea of the PAC data structure is then to
extend subgoal frames with an auxiliary private chaining of answers
for each subgoal call, in order to keep track of the answers already
found for the call. Later, if a thread completes a subgoal's
evaluation, i.e, if the subgoal's table is marked as complete, its PAC
is made public, so that from that point on all threads can use that
chain in complete (only reading) mode. Figure~\ref{fig_tabtries_pcc}
illustrates the new data structures involved in the implementation of
our PAC's proposal for a situation where several threads are
evaluating the same tabled subgoal call $call\_i$.

@@

Figure~\ref{fig_tabtries_pcc}(a) shows then a situation where two
threads, $T_1$ and $T_{k-1}$, are sharing the same subgoal entry for
the call $call\_i$ when the subgoal is still under evaluation, i.e.,
the subgoal is not yet complete. The current state of the evaluation
shows an answer trie with 3 answers found for the subgoal call
$call\_i$. For the sake of simplicity, we are omitting the internal
answer trie nodes and we are only showing the leaf nodes nodes $LN_1$,
$LN_2$ and $LN_3$, in the figure.

@@

By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
thread, on the minimum overhead ratio, local scheduling is better on
the NS design, while batched scheduling is better in the FS
design. For the average and maximum overhead ratios, in the FS design,
the best strategy is always the local scheduling. Additionally, the
difference between local and batched scheduling in the FS is higher
than in the NS design, which indicates the overhead that the PAC
strategy brings to the FS design. Recall, whenever an answer is found
during the evaluation, PAC requires threads to traverse their private
consumer data structures to check if the answer was already
propagated. This is the reason for difference in the overheads that we
are observing in the experimental results. As we scale the number of
threads, for the NS design both scheduling strategies have similar
minimum, average and maximum overhead ratios. For the FS design, the
best minimum overhead ratio is always for the batched scheduling. For
the average and maximum overhead ratio, local scheduling is always
better than batched scheduling. However, for the average ratio, the
results between both strategies are quite close. For the maximum
overhead ratio, the difference between both strategies is
significantly higher than in the NS design. In conclusion,
experimental results show that on average, the PAC strategy does not
seem to have a big impact in the performance, however it still leaves
room for further improvements. The difference between both local and
batched scheduling is significantly higher in the FS design than in
the NS design.

@@

Local and batched scheduling are arguably two of the most well-known
tabling scheduling techniques. The major difference between them is
that local propagates answers only after all answer are found, while
batched propagates answers immediately after they are found. Batched
scheduling is a useful strategy in tabled logic programs that require
an eager propagation of answers and/or do not require the complete set
of answers to be found (partial evaluation of the search space). In
this work, we have presented the PAC strategy, which is a simple and
novel approach for combining the FS design with batched
scheduling. PAC splits answer representation from answer propagation,
and allows the first to be publicly shared among threads while the
second to be private to each thread.

>>>>

\begin{abstract}
Tabling is a technique that overcomes some limitations of traditional
Prolog systems in dealing with redundant sub-computations and
recursion. Multithreading in Prolog is the ability to concurrently
perform computations, in which each computation runs independently but
shares the program clauses. When multithreading is combined with
tabling, we have the best of both worlds, since we can exploit the
combination of higher procedural control with higher declarative
semantics. To support this combination, the Yap Prolog system has, at
engine level, multiple designs that vary from a No-Sharing design,
where each thread allocates fully private tables for each new tabled
subgoal called during its computation, to a Full-Sharing design, where
threads share the complete table space. Arguably, batched scheduling
is one of the most successful tabling scheduling strategies. In this
work, we propose the PAC strategy to combine the Full-Sharing design
with batched scheduling. Experimental results show that, PAC is a good
first approach, since with it the FS design remains quite
competitive.\\

@@

Tabling~\cite{Chen-96} is a technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Tabling consists in storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of
XSB. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

Multithreading in Prolog is the ability to concurrently perform
computations, in which each computation runs independently but shares
the program clauses. When multithreading is combined with tabling, we
have the best of both worlds, since we can exploit the combination of
higher procedural control with higher declarative semantics. To the
best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
are the only Prolog systems that support the combination of
multithreading with tabling. Yap implements a SWI-Prolog compatible
multithreading library~\cite{Wielemaker-03}. For multithreaded tabled
evaluation, a thread views its tables as private but, at the engine
level, Yap has three designs~\cite{Areias-12a}. These designs vary
from a \emph{No-Sharing} (NS) design, where each thread allocates
private tables for each new tabled subgoal called during its
evaluation, to a \emph{Full-Sharing} (FS) design, where threads share
the complete table space.

The decision about the evaluation flow is determined by the
\emph{scheduling strategy}. Different strategies may have a
significant impact on performance, and may lead to a different
ordering of solutions to the query goal. Arguably, the two most
successful tabling scheduling strategies are \emph{local scheduling}
and \emph{batched scheduling}~\cite{Freire-96}. Local scheduling tries
to complete subgoals as soon as possible. When new answers are found,
they are added to the table space and the evaluation fails. Local
scheduling has the advantage of minimizing the size of clusters of
dependent subgoals, however it delays propagation of answers and
requires the complete evaluation of the search space.

@@

In this work, we propose an extension to the table space data
structures, which we named \emph{Private Answer Chaining (PAC)}, as a
way to keep track, per thread and subgoal call, of the answers that
were already found and already propagated. We discuss in detail our
proposal for extending the FS design with batched scheduling and we
present a performance analysis comparison between local and batched
scheduling. Experimental results show that, despite the extra PAC data
structures required, the execution time is still quite competitive.

The remainder of the paper is organized as follows. First, we briefly
introduce some background and related work. Then, we describe our PAC
approach and we discuss the most important implementation
details. Finally, we present experimental results and we end by
outlining some conclusions.

@@

\begin{figure}[!ht]
\vspace{-0.5\intextsep}
\centering
\includegraphics[width=9.5cm]{figures/yap-mt.pdf}
\caption{Yap's multithreaded table space organization for the NS and FS designs}
\label{fig_yap_mt_support}
\vspace{-0.5\intextsep}
\end{figure}

@@

On the other hand, batched scheduling evaluates a tabled logic program
in a depth-first manner. It favors the forward execution first instead
of backtracking, leaving the consumption of answers and completion for
last. It thus tries to delay the need to move around the search tree
by batching the return of answers. When new answers are found for a
particular tabled subgoal, they are added to the table space and the
execution continues. For some situations, this results in creating
dependencies to older subgoals, therefore enlarging the current CDS
and delaying the fix-point that guarantees that all dependent subgoals
in a CDS are completely evaluated~\cite{Sagonas-98}. Batched
scheduling can be an useful strategy in tabled logic programs that
require an eager propagation of answers and/or do not require the
complete set of answers to be found.

@@

\begin{wrapfigure}{r}{6.5cm}
\vspace{-0.5\intextsep}
\centering
\includegraphics[width=6.5cm]{figures/pcc_key-idea.pdf}
\caption{PAC splits answer representation and answer propagation}
\label{fig_tabtries_pcc_key-idea}
\vspace{-0.5\intextsep}
\end{wrapfigure}

In this section, we describe our proposal to support the combination
of batched scheduling with the FS design. In the FS design answer
representation and answer propagation is stored in the answer trie
data structure, thus threads are unable to know if they have or not
have propagated an answer whenever it appears during the
evaluation. To solve that, we propose an extension to the table space
data structures, which we named \emph{Private Answer Chaining} (PAC),
as a way to keep track, per thread and subgoal call, of the answers
that were already found and propagated to the thread's repeated
calls. Figure~\ref{fig_tabtries_pcc_key-idea} illustrates the key idea
of the PAC strategy. In a nutshell, PAC splits answer representation
from answer propagation, and allows the first to be kept publicly
shared among threads in the answer trie data structure, and the second
to be privately stored in the subgoal frame data structure of each
thread.

@@

The PAC procedure works at the subgoal frame level. The key idea of
PAC is to extend subgoal frames with an auxiliary private chaining of
answers for each subgoal call, in order to keep track of the answers
already found for the call. Later, if a thread completes a subgoal's
evaluation, i.e, if the subgoal's table is marked as complete, its PAC
is made public, so that from that point on all threads can use that
chain in complete (only reading) mode. Figure~\ref{fig_tabtries_pcc}
illustrates the new data structures involved in the implementation of
our PAC's proposal for a situation where several threads are
evaluating the same tabled subgoal call $call\_i$.

@@

Figure~\ref{fig_tabtries_pcc}(a) shows then a situation where two
threads, $T_1$ and $T_{k-2}$, are sharing the same subgoal entry for
the call $call\_i$ when the subgoal is still under evaluation, i.e.,
the subgoal is not yet complete. The current state of the evaluation
shows an answer trie with 3 answers found for the subgoal call
$call\_i$. For the sake of simplicity, we are omitting the internal
answer trie nodes and we are only showing the leaf nodes nodes $LN_1$,
$LN_2$ and $LN_3$, in the figure.

@@

By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
thread, on the minimum overhead ratio, local scheduling is better on
the NS design, while batched scheduling is better in the FS
design. For the average and maximum overhead ratios, in the FS design,
the best strategy is always the local scheduling. Additionally, the
difference between local and batched scheduling in the FS is higher
than in the NS design, which indicates the overhead that the PAC
strategy brings to the FS design. Recall, whenever an answer is found
during the evaluation, PAC requires threads to traverse their private
consumer data structures to check if the answer was already
propagated. This is the reason for difference in the overheads that we
are observing in the experimental results. As we scale the number of
threads, for the NS design both scheduling strategies have similar
minimum, average and maximum overhead ratios. For the FS design, the
best minimum overhead ratio is always for the batched scheduling. For
the average and maximum overhead ratio, local scheduling is always
better than batched scheduling. However, for the average ratio, the
results between both strategies are quite close. For the maximum
overhead ratio, the difference between both strategies is
significantly higher than in the NS design. In conclusion,
experimental results show that on average, the PAC strategy does not
seem to have a big impact in the performance, however it still leaves
room for further improvements, since the difference between local and
batched scheduling is higher in the FS design than in the NS design.

@@

Local and batched scheduling are arguably two of the most well-known
tabling scheduling techniques. The major difference between them is
that local propagates answers only after all answer are found, while
batched propagates answers immediately after they are found. Batched
scheduling is a useful strategy in tabled logic programs that require
an eager propagation of answers and/or do not require the complete set
of answers to be found. In this work, we have presented the PAC
strategy, which is a simple and novel approach for combining the FS
design with batched scheduling. PAC splits answer representation from
answer propagation, and allows the first to be publicly shared among
threads while the second to be private to each thread.
